---
title: "ANES2024 Analysis"
output:
  pdf_document: default
  html_document: default
---

**Name:** Morgan Powers<br>
**Research Question:** What factors influence voters to support a candidate who does not align with their partisan ideology?<br>
**Hypothesis:** I suspect that a combination of economic status, education, foreign policy views, and concern about crime rates and immigration will have high correlations.<br>
**Project GitHub repository:** https://github.com/Scimitaria/ResearchMethods-FinalProject/<br>
<small>Note on GitHub usage for Dr. Bussing: Only a few of the features of GitHub are relevant to this project. The link itself takes you to a directory with all of the project files. Clicking on one of the files takes you to the code/data, with an opportunity to download the file. If for whatever reason you want to view past work in the repository, you can click ‘<x> commits’ in the top right corner, select a commit, and press ‘browse files’.</small>
<br>Documentation of variable declarations is in data.r.<br><br>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
source("data.r")
```

#### Candidate thermometer ratings
```{r}
printf("mean Harris thermometer rating",mean(anes$harrisTherm,na.rm=T))
printf("mean Trump thermometer rating",mean(anes$trumpTherm,na.rm=T))
```

#### Economic Worry
```{r}
prints("Voters who consider themselves liberal but voted for Trump from most to least worried about their financial situation.")
print(table(anes$leftTheLeft,anes$worry))
prints("Voters who consider themselves conservative but voted for Harris from most to least worried about their financial situation.")
print(table(anes$renouncedTheRight,anes$worry))
prints("Voters who voted against their ideological bias from most to least worried about their financial situation.")
print(table(anes$switch,anes$worry))
prints("Correlation of the count of switch voters and financial worry level")
#lm(switch ~ worry,data=anes)
cor.test(anes$worry,anes$switch,use="complete.obs")
```
Voters’ level of worry about their financial situation (ANES$V241539) displays a statistically significant inverse correlation with status as a switch voter (p-value of 0.0224, coefficient estimate of 0.0078, bearing in mind that higher values of V241539 represent lower levels of worry).
In other words, people who feel secure in their finances were slightly more likely to be switch voters.
However, as evidenced by the low correlation coefficient, the trend is not very strong.
<br>
#### Importance of Crime
```{r}
prints("Voters who consider themselves liberal but voted for Trump from most to least worried about crime.")
print(table(anes$leftTheLeft,anes$crime))
prints("Voters who consider themselves conservative but voted for Harris from most to least worried about crime.")
print(table(anes$renouncedTheRight,anes$crime))
prints("Voters who voted against their ideological bias from most to least worried about crime.")
print(table(anes$switch,anes$crime))
prints("Correlation of the count of switch voters and importance of crime")
cor.test(anes$crime,anes$switch,use="complete.obs")
```
While this data does display a negative correlation (switch voters placing more importance on crime), the p-value is 0.43, so the correlation is not significant.
